Lens Modeling in \sw

### 2014-06-30 text/partA.tex
more edits

An implementation (\spl) is tested in a modelling challenge using 29
simulated lenses out of a larger set created for a citizen-science
lens search (\sw).  We find that amateur lensers inferred the image
parities and time ordering consistently in some lenses, while tending
to make errors in other lenses, depending on the image morphology.
Errors in image parity and time ordering lead to large errors in the
mass distribution, but the enclosed mass is more robust.  Shortcomings
include (a)~lens profiles that are too shallow, (b)~over-estimation of
the Einstein radius by $\sim25\%$, (c)~lack of detailed source
modelling.  Ideas for improvement are discussed.
\end{abstract}

@@

\begin{figure}
  \centering
  \subfigure{\includegraphics[width=0.8\linewidth]
             {fig/007025_kappa_encl}}
  \subfigure{\includegraphics[width=0.8\linewidth]
             {fig/007022_kappa_encl}}
  \subfigure{\includegraphics[width=0.8\linewidth]
             {fig/007021_kappa_encl}} \\
  \caption{Enclosed mass \kenc from three models of the same system.
    The upper panel corresponds to the model shown in \figref{7025},
    with the image-parities incorrect by a quarter-turn.  The middle
    corresponds to \figref{7022}, where the image parities were
    correct.  The lower panel corresponds to another model, where the
    image parities were incorrect by a half-turn.}
\label{fig:kapenc_compare_faulty}
\end{figure}

Figure \ref{fig:kapenc_compare_faulty} shows a comparison of input and
recovered mass profiles.  The panel shows a average $\kappa$ within a
given radius, as a function of radius.  The red curve is the true
value, and where it crosses unity (dotted horizontal line) is the
notional Einstein radius $\Theta_{\text{E, sim}}$.  The two blue
curves are the minimal and maximal mean enclosed $\kappa$ from the
internal ensemble in \spl.  The region between the blue curves is
shaded between the radii of the innermost and the outermost images ---
this is the confidence region from the modeling.  As we see, the
shaded blue is slightly above the red curve.  The Einstein Radius
$\Theta_\text{E}$ of the model is estimated crossing the mean enclosed
$\kappa$ (not plotted) with unity.

The panel shows a average $\kappa$ within a given radius, as a
function of radius.  The red curve is the true value, and where it
crosses unity (dotted horizontal line) is the notional Einstein radius
$\Theta_{\text{E, sim}}$.  The two blue curves are the minimal and
maximal mean enclosed $\kappa$ from the internal ensemble in \spl.
The region between the blue curves is shaded between the radii of the
innermost and the outermost images --- this is the confidence region
from the modeling.  As we see, the shaded blue is slightly above the
red curve.  The Einstein Radius $\Theta_\text{E}$ of the model is
estimated crossing the mean enclosed $\kappa$ (not plotted) with
unity. \citep[see, e.g.,][]{2002LNP...608....1C}.

>>>>

An implementation (\spl) is tested in a modelling challenge using 29
simulated lenses out of a larger set created for a citizen-science
lens search (\sw).  We find that amateur lensers inferred the image
parities and time ordering consistently in some lenses, while tending
to make errors in other lenses, depending on the image morphology.
Errors in image parity and time ordering lead to large errors in the
mass distribution, but the enclosed mass is more robust.  Shortcomings
include (a)~lens profiles that tend to be systematically shallower
than the lenses, and consequent over-estimation of the Einstein radius
by $\sim25\%$, (c)~lack of detailed source modelling.  Ideas for
improvement are discussed.
\end{abstract}

@@

### 2014-06-30 text/partB.tex
more edits

\tabref{stats} presents a summary of the test.  The evaluation was
done manually, comparing the input to \spl with the actual
arrival-time surface of the sim.  This amounts to comparing the
upper-left and middle-left panels in each of
Figures~\ref{fig:6941}--\ref{fig:7022}, and similaly the other 113
models.

\begin{table}\centering\begin{tabular}{llcc}\hline
     & Total & 119 \\
\hline
 R1: & images approximately on correct locations & 110 \\
 R2: & images parities and time-ordering correct & 70 \\
\hline
 E01: & inaccurate image placement over an arc & 21 \\
 E02: & swapped minimum and saddle in double  & 2 \\
 E03: & identified two images of four & 5 \\
 E04: & modelled a three-image arc as one image & 4 \\
 E05: & swapped early and late saddles in a quad & 7 \\
 E06: & swapped minima and saddles in quad & 38 \\
 E07: & missed faint images & 1 \\
 E08: & modelled one image as a three-image arc & 5 \\
 E09: & identified two nearby images as one & 3 \\
 E10: & proposed too many images & 1 \\
\hline
\end{tabular}
\caption{Table of image-identification errors and the number of models
  containing each.}
\label{tab:stats}
\end{table}

The images of the system are considered to be identified correctly, if all the images have been identified and are approximated within $\pm0.05\cdot\text{imgage width}$.
The parity is considered correct, if those identified points have the right ordering with respect to arrival time.

Additionally, ten types of errors (labeled E01 -- E10, listed in \tabref{stats}) that occurred were identified.
Each generated model could contain more than one error.

We conclude that the volunteers are performing very well identifying and positioning images, with a performance of 92\% (R1, p=0.92).
Most of the problems where due to unclear arc-like structures (E01, p=0.18; E04, p=0.03; E08, p=0.04).
Critical errors like the failure to identify all five images in a five images system (E03, p=0.04) or to include too many images (E10, p=0.01) did almost never happen.
From this we conclude that the introduction materials was adequate and the volunteers understand the basics of gravitational lensing.

The assignment of the parity of the images was a more difficult task.
In 59\% (R2, p=0.59, N=70) of the cases the volunteers succeeded to identify the right configuration.
Most of the failures are due to E06 (N=38, p=32\%), followed by E05 (N=7, p=6\%).
E06 describes a situation, where the minima and saddle points of a five image configuration were exchanged (rotated by $\pm90\dgr$), see \figref{7022} for an example.
E05 describes a situation, where the ordering of the saddle points was wrong (rotation by $180\dgr$).
While these errors occurred, we suspect they can be avoided with better training material and some examples for the obvious cases.
For more challenging cases, like very symmetrical distribution of the lensed images (for example model 7022, \figref{7022}), those errors should still produce plausible results, as will be explained in the next section.

@@

The second test was to compare the mass distributions of the lens $\kappa(x, y)$ given by the sims and generated by the volunteers.
To get a means of comparing the sims to the models, the total convergence\footnote{often called enclosed mass} $\kappa_{\text{encl}}(r)$ was calculated for both.
The Einstein radius $\Theta_\text{E}$ is defined by $\kappa_{\text{encl}}(\Theta_\text{E})=1$ and gives a number that allows a rough comparison between a sim and a model.
We also let an expert (PS) model three selected systems to compare the results from volunteers to those of a professional.

To compare the enclosed mass profile and the Einstein radius of the simulation and the models, \kenc was calculated using the mass map \kap[x,y] generated in the modeling process.
From the ensemble of models generated by one modeling process, the mean is taken as the resulting $\kappa(r)$ to calculate \ERg.
To estimate the errors, the extremal models are used to estimate a lower and upper limit for \ERg.
These results can be seen in \figref{ER_per_sim}.
This figure shows that this technique of estimating the error using the ensemble of all models underestimates the error significantly and should be improved for further analysis.

\begin{figure}
  \centering
  \subfigure{\includegraphics[width=0.9\linewidth]
             {fig/007022_kappa_encl}}
  \subfigure{\includegraphics[width=0.9\linewidth]
             {fig/007025_kappa_encl}}
  \subfigure{\includegraphics[width=0.9\linewidth]
             {fig/007021_kappa_encl}} \\
  \caption{
  %Enclosed mass \kenc from three models of a system with incorrect image identifications. \figref{7022} shows a model (by an expert) with the correct identifications, while \figref{7025} shows a model with error E06 (parities rotated by $\pi/2$).  Shown here are the \kenc results for two models with error E06 and one model (bottom) with error E05 (parities rotated by $\pi$).
  }
\label{fig:kapenc_compare_faulty}
\end{figure}

\begin{figure*}
  \centering
    \includegraphics[width=0.90\linewidth]{fig/eR_5}
  \caption{Relative \ERf / \ERg[, sim] for models by volunteers (blue cross), models made by an expert (red cross with offset), including rejected models (green squares); binned per sim.}
  \label{fig:ER_per_sim}
\end{figure*}

\Figref{ER_per_sim} shows that the calculated \ERf of the models tend
to be too high.  The overshoot varies from around 0.2 to 0.4 for good
models.  Discounting the models which were flagged by the users as
poor, \ERf{E,rec} is overestimated.  The mean and standard deviation
of the overestimate are 25\% and 20\%.  The models by an expert also
show a similar bias, having mean and standard deviation of the
overestimate being 20\% and 8\%.
One of the reasons for this is that it is hard to get the center of the lens on spot.
An offset leads to a flatter mass profile for the model compared to the simulation.

Comparing the models from volunteers and experts can be done in Figures \ref{fig:7022}, \ref{fig:7025} and \ref{fig:kapenc_compare_faulty}, where only the
expert got the right configuration (\figref{7022}) but enclosed masses are all fairly similar.  In cases with nearly symmetric image configurations, like this example, it is more difficult to identify the image parities correctly.  Incorrect image-parity identification changes the orientation of the mass distribution \kap[x,y],
but \kenc and thus the \ERf are not influenced that much.

Two additional sims were modeled by an expert, \asw{1hpf} and \asw{0vqg}.
Looking at the results for \ERg for those models in \figref{ER_per_sim}, we conclude that the performance of volunteers (blue crosses) and experts (red crosses, offset) is comparable.
%Note that the models of \asw{0vqg} with \ERg[, rel] around 0.25 (6935 -- 6937) are failed models that show the attempts of a single user, that came finally up with model 6938 as final result.\todo{remove?}

\clearpage
\section{Outlook} \label{sec:todo}

The lens-modeling challenge indicates that the Spacewarps collective
is good, not only for identifying lens candidates for follow-up, but
modeling candidates as well.

>>>>

\tabref{stats} presents a summary of the test.  The evaluation was
done manually, comparing the input to \spl with the actual
arrival-time surface of the sim.  This amounts to comparing the
upper-left and middle-left panels in each of
Figures~\ref{fig:6941}--\ref{fig:7022}, and similarly the other 111
models.

\begin{table}\centering\begin{tabular}{lr}\hline
    Total & 119 \\
\hline
    errors in image locations  & 9 \\
    errors in image parities or time ordering & 49 \\
\hline
    inaccurate image placement over an arc & 21 \\
    identified two images of four & 5 \\
    identified two nearby images as one & 3 \\
    missed faint images & 1 \\
    proposed too many images & 1 \\
    modelled a three-image arc as one image & 4 \\
    modelled one image as a three-image arc & 5 \\
    swapped minimum and saddle in double  & 2 \\
    swapped minima and saddles in quad & 38 \\
    swapped early and late saddles in a quad & 7 \\
\hline
\end{tabular}
\caption{Table of image-identification errors and the number of models
  containing each.  A model can contain more than one type of error.}
\label{tab:stats}
\end{table}

The images of the system are considered to have been located
correctly, if all the images have been identified and are approximated
within about 5\% of \spl frame used to draw the spaghetti diagram.
That frame size is adjustable by the user, but in practice it is
somewhat larger than the spaghetti diagram.  Such image-placement
errors were found in only 9 models.  That does not include inaccurate
image placement over an arc, which was considered a separate category
of error.

In addition to simple image-placement errors, ten types of errors were
recognised and are listed in \tabref{stats}.  Most of the problems
where due to unclear arc-like structures.  Critical errors like the
failure to identify all five images in a five images system, or to
include too many images, were rare.

The assignment of the parity of the images was a more difficult task,
and was successful in only 60\% of the cases.  The most common error
was swapping of minima and saddle-points in a quad; \figref{7025}
shows an example.  Another, less common, error was flipping the
spaghetti diagram, this swapping the time-ordering of the two saddle
points.

Incorrect image parities and time orderings tended to produce
poorer-looking models, such as the checkboard patterns in the mass map
in \figref{7025}.  Interestingly, however, the enclosed-mass profiles
were quite robust.  We will consider this aspect in the next section.

\FloatBarrier

\begin{figure}
  \centering
  \subfigure{\includegraphics[width=0.8\linewidth]
             {fig/007025_kappa_encl}}
  \subfigure{\includegraphics[width=0.8\linewidth]
             {fig/007022_kappa_encl}}
  \subfigure{\includegraphics[width=0.8\linewidth]
             {fig/007021_kappa_encl}} \\
  \caption{Mean $\kappa$ inside a circle around the lens centre, as a
    function of the radius of the circle.  (See \secref{tests.t2} for
    details.) The upper panel corresponds to the model shown in
    \figref{7025}, in which the minima and saddle-points have been
    incorrectly swapped.  The middle corresponds to \figref{7022},
    where the image parities were correct.  The lower panel
    corresponds to another model, where the image parities were
    correct but the time-ordering was incorrect.}
\label{fig:kapenc_compare_faulty}
\end{figure}

@@

The second test was to compare the mass distributions $\kappa(x,y)$ of
the sims and of the \spl models.  A visual comparison is presented for
the eight models in Figures~\ref{fig:6941}--\ref{fig:7022}, in the
lower-left versus lower-right panels.  We will summarize the mass
distributions drastically in a single number, the effective Einstein
radius.  Other measures for comparison of free-form lensing mass
distributions appear in \cite{2014arXiv1401.7990C}, but comparing
Einstein radii is already useful.

There is no standard way of defining the Einstein radius of a general
non-circular lens.  We adopt the simple definition
\begin{equation} \label{eq:effRE}
 \langle\kappa\rangle_{\Theta_E} = 1,
\end{equation}
that is, the effective Einstein radius \ERf is such that the
mean $\kappa$ is unity inside a circle of radius \ERf centred at
the lens centre.

To illustrate, \figref{kapenc_compare_faulty} compares the circularly
averaged mass profiles three different models of one particular lens;
two of the models are shown in Figures~\ref{fig:7025} and
\ref{fig:7022}. Each panel in \figref{kapenc_compare_faulty} shows the
mean $\kappa$ within a circle of given radius.  The red curve is the
correct profile for the sim.  The two blue curves are the minimal and
maximal mean enclosed $\kappa$ from the internal ensemble in \spl.
Radial locations of the images are marked, along with the image
parities.  The region between the blue curves is shaded between the
radii of the innermost and the outermost images: this is the
confidence region from the modelling.  The definition (\ref{eq:effRE})
for \ERf corresponds to crossing the dashed horizontal line at
$10^0$: the red curve crosses the dashed line at the actual Einstein
radius \ERf[,act]; the recovered Einstein radius
\ERf[,rec] and its uncertainty are given by the blue curves
crossing the dashed line.  We see that in all three panels, the blue
curves are shallower than the red curve and \ERf[,rec] is
more than \ERf[,act], by more than the model uncertainties.
Now, steeper mass profiles tend to give wider image separations ---
recall that the image separation for a circular isothermal lens is
$2\Theta_E$, whereas for a point mass it is more \citep[see,
  e.g.,][]{2002LNP...608....1C} --- so \ERf[,rec] being too
high is really a consequence of the \spl models being too shallow for
the sims.

\begin{figure*}
  \centering
    \includegraphics[width=0.90\linewidth]{fig/eR_5}
  \caption{Model-recovered versus actual Einstein radii \ERf[,rec] and
    \ERf[,act].  Circles denote models where the image parities and
    time-orderings were correct, with filled circles for models by an
    expert.  Crosses mark models where the time-ordering was
    incorrect.  Plus signs indicate models flagged by the modeller as
    failures.}
  \label{fig:ER_per_sim}
\end{figure*}

\Figref{ER_per_sim} shows that \ERf[,rec] of the models tend
to be too high.  Discounting the models which were flagged by the
users as poor, the mean and standard deviation of the overestimate are
25\% and 20\%.  The models by an expert (PS) also show a similar bias,
having mean and standard deviation of the overestimate being 20\% and
8\%.  One of the reasons for this is that it is hard to get the center
of the lens on spot.  An offset leads to a flatter mass profile for
the model compared to the simulation.

\section{Outlook} \label{sec:todo}

The lens-modeling challenge indicates that the Spacewarps collective
is good, not only for identifying lens candidates for follow-up, but
modelling candidates as well.
